Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-02-05 paper.tex
edits in grammar induction section

The recognition model $q$ is a neural network that predicts,
for each task $x\in X$, a weight vector $q(x) = \theta^{(x)}\in \mathbb{R}^{|\mathcal{D}| + 1}$.
Together with the DSL, this defines a distribution over programs,
$\probability[p|\mathcal{D},\theta = q(x)]$.
We abbreviate this distribution as $q(p|x)$.
The crucial aspect of this framing is that the neural network
leverages the structure of the learned DSL,
and so is \emph{not} responsible for
generating programs wholesale.

We want a recognition model that closely approximates the true posteriors over programs, and so aim to minimize the following KL-divergence:
\begin{equation*}
  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]||q(p|x) \right) \right]
\end{equation*}
which is equivalent to maximizing
$$
  \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]
$$
  where the expectation is taken over tasks. One could take this expectation
  over the empirical distribution of the observations,
  like how an autoencoder is trained~\cite{hinton2006reducing}; or, one could take this expectation over samples from the generative model, like how a Helmholtz machine is trained~\cite{dayan1995helmholtz}.
  We found it useful to maximize both an autoencoder-style objective (written $\mathcal{L}_{\text{AE}}$) and a Helmholtz-style objective ($\mathcal{L}_{\text{HM}}$), giving the \system objective for a recognition model, $\mathcal{L}_{\text{RM}}$:
  \begin{align}
  \mathcal{L}_{\text{RM}}& = \mathcal{L}_\text{AE} + \mathcal{L}_\text{HM}\\
  \mathcal{L}_{\text{HM}}& = \expect_{p\sim(\mathcal{D},\theta) }\left[\log q(p|x)\right],\text{ $p$ evaluates to $x$}\nonumber\\
  \mathcal{L}_{\text{AE}}& = \expect_{x\sim X}\left[\sum_{p\in \mathcal{F}_x}
    \frac{\probability\left[x,p|\mathcal{D},\theta \right]}{\sum_{p'\in \mathcal{F}_x}\probability\left[x,p|\mathcal{D},\theta \right]}\log q(p|x)\right]\nonumber
  \end{align}

@@

Exact maximization of $J$ (Eq.\ref{lowerBound}) w.r.t. $(\mathcal{D},\theta)$
is intractable,
so we take a more heuristic approach.
The strategy is to search locally through the space of DSLs,
proposing small local changes to $\mathcal{D}$ until $J$ fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting subexpressions from
programs already in the frontier.
These extracted subexpressions
are fragments of the original programs, and can introduce new variables (Figure~\ref{fragmentExample}),
which then become new functions in the DSL.

Closely related to Fragment Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing},
but context-sensitive

@@

Additionally, for each proposed $\mathcal{D}$ we have to reestimate $\theta$.
Although this problem may seem 
very similar to estimating the parameters of a probabilistic context free grammar (PCFG: see~\cite{}),
for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
$\mathcal{D}$
 is actually context-sensitive
due to the presence of variables in the programs and also due to the
polymorphic typing system.
In the Appendix we derive a tractable MAP estimator for $\theta$.
\begin{algorithm}[tb]
   \caption{DSL Induction Algorithm}
   \label{grammarInductionAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
     \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
     \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
     \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
     \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - ||\theta||_0$
     \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
     \WHILE {true}
     \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ subexpression of }z\}$
     \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
     \STATE \textbf{if }$\text{score}(\mathcal{D}') < \text{score}(\mathcal{D})$\textbf{ return }$\mathcal{D},\theta^*(\mathcal{D})$
     \STATE $\mathcal{D}\gets\mathcal{D}'$
     \ENDWHILE
   \end{algorithmic}
\end{algorithm}

@@

Algorithm~\ref{mainAlgorithm} describes how we combine the
program search, recognition model training, and DSL induction.
We additionally make the following implementation details:
(1) On the first iteration we do \emph{not} train the recognition model on samples from
the generative model because, on the first iteration, we have not yet learned a
new DSL, so we instead trained the network to only maximize $\mathcal{L}_{\text{AE}}$;
(2) Because the frontiers can grow very large,
we only keep around the top $10^4$ programs $p$ in each frontier $\mathcal{F}_x$
with the highest $\probability[x,p|\mathcal{D},\theta]$;
(3) During both DSL induction and neural net training, we calculate $\text{score}(\mathcal{D}')$ and $\mathcal{L}_{\text{RM}}$
by only summing over the top $K$
programs in $\mathcal{F}_x$.
We found that $K = 2$ sufficed.
(4) For added robustness, at each iteration we enumerate programs from both the generative model and the recognition model.

>>>>

The recognition model $q$ is a neural network that predicts,
for each task $x\in X$, a weight vector $q(x) = \theta^{(x)}\in \mathbb{R}^{|\mathcal{D}| + 1}$.
Together with the DSL, this defines a distribution over programs,
$\probability[p|\mathcal{D},\theta = q(x)]$.
We abbreviate this distribution as $q(p|x)$.
The crucial aspect of this framing is that the neural network
leverages the structure of the learned DSL,
so it is \emph{not} responsible for
generating programs wholesale.

We want a recognition model that closely approximates the true posteriors over programs. We formulate this as minimizing the expected KL-divergence:
\begin{equation*}
  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]||q(p|x) \right) \right]
\end{equation*}
which is equivalent to maximizing
\begin{equation*}
  \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]
\end{equation*}
where the expectation is taken over tasks. One could take this expectation
over the empirical distribution of the observations,
like how an autoencoder is trained~\cite{hinton2006reducing}; or, one could take this expectation over samples from the generative model, like how a Helmholtz machine is trained~\cite{dayan1995helmholtz}.
We found it useful to maximize both an autoencoder-style objective (written $\mathcal{L}_{\text{AE}}$) and a Helmholtz-style objective ($\mathcal{L}_{\text{HM}}$), giving the \system objective for a recognition model, $\mathcal{L}_{\text{RM}}$:
\begin{align}
\mathcal{L}_{\text{RM}}& = \mathcal{L}_\text{AE} + \mathcal{L}_\text{HM}\\
\mathcal{L}_{\text{HM}}& = \expect_{p\sim(\mathcal{D},\theta) }\left[\log q(p|x)\right],\text{ $p$ evaluates to $x$}\nonumber\\
\mathcal{L}_{\text{AE}}& = \expect_{x\sim X}\left[\sum_{p\in \mathcal{F}_x}
  \frac{\probability\left[x,p|\mathcal{D},\theta \right]}{\sum_{p'\in \mathcal{F}_x}\probability\left[x,p|\mathcal{D},\theta \right]}\log q(p|x)\right]\nonumber
\end{align}

@@

Exact maximization of $J$ (Eq.\ref{lowerBound}) w.r.t. $(\mathcal{D},\theta)$
is intractable,
so we take a more heuristic approach.
The strategy is to search locally through the space of DSLs,
proposing small local changes to $\mathcal{D}$ until $J$ fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting subexpressions from
programs already in the frontier.
These extracted subexpressions
are fragments of the original programs, and can introduce new variables (Figure~\ref{fragmentExample}),
which then become new functions in the DSL.
Our model of subexpressions is loosely related to Fragment
Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing},
but context-sensitive.

@@

To appropriately score each proposed $\mathcal{D}$ we must reestimate the weight vector $\theta$.
Although this problem may seem 
very similar to estimating the parameters of a probabilistic context free grammar (PCFG: see~\cite{}),
for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
a DSL in \system may be context-sensitive due to the presence of variables
in the programs and also due to the polymorphic typing system.
In the Appendix we derive a tractable MAP estimator for $\theta$.
\begin{algorithm}[tb]
   \caption{DSL Induction Algorithm}
   \label{grammarInductionAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
     \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
     \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
     \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
     \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - ||\theta||_0$
     \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
     \WHILE {true}
     \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ subexpression of }z\}$
     \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
     \STATE \textbf{if }$\text{score}(\mathcal{D}') < \text{score}(\mathcal{D})$\textbf{ return }$\mathcal{D},\theta^*(\mathcal{D})$
     \STATE $\mathcal{D}\gets\mathcal{D}'$
     \ENDWHILE
   \end{algorithmic}
\end{algorithm}

@@

Algorithm~\ref{mainAlgorithm} describes how we combine program search,
recognition model training, and DSL induction.
We note the following implementation details:
(1) on the first iteration, we do \emph{not} train the recognition model on samples from
the generative model because a DSL has not yet been learned --
we instead train the network to only maximize $\mathcal{L}_{\text{AE}}$.
(2) Because the frontiers can grow very large,
we only keep around the top $10^4$ programs $p$ in each frontier $\mathcal{F}_x$
with the highest likelihood $\probability[x,p|\mathcal{D},\theta]$.
(3) During both DSL induction and neural net training, we calculate $\text{score}(\mathcal{D}')$ and $\mathcal{L}_{\text{RM}}$
by only summing over the top $K$
programs in $\mathcal{F}_x$ -- we found that $K = 2$ sufficed.
(4) For added robustness, at each iteration we enumerate programs from both the generative model and the recognition model.
